Rapid shortening at the eastern margin of the Tibetan plateau prior to the 2008 Mw=7.9 Wenchuan earthquake

### 2014-09-12 paper.tex
Deleted crazy number of line breaks in Abstract.
Going to add the second figure. Working on the results description.

\title{Rapid shortening at the eastern margin of the Tibetan plateau prior to the 2008 Mw=7.9 Wenchuan earthquake}
\author{T. Ben Thompson, Brendan J. Meade}
\date{September 2014}
\usepackage{graphicx}
\usepackage[margin=1.0in]{geometry}

@@

However, it remains to be explained how slow present-day geodetic shortening estimates can be reconciled with the Wenchuan earthquake and the large structurally-observed shortening.
Prior estimates can be categorized into those that neglect earthquake cycle effects, analyzing the fault slip-deficit without no explicit model of the geometry or tectonics, (CHEN, SHEN) and those that use block models, which normally assume a simplified fault geometry (CITE MEADE 2007, BURCHFIEL 2008, LOVELESS AND MEADE 2011).
Loveless and Meade report the highest shortening rate of 3.2 mm/yr.
Here, we present a model that includes earthquake cycle effects to explain the regional geodetic velocities as the result of a complex interseismically locked fault geometry, including a 20km deep detachment and the range-front thrust Beichuan fault.
This geometry is consistent with structural interpretations and with the geometry of Qi et al 2012 and Fielding et al 2013. 
Using this range-front thrust plus detachment geometry, we infer an average horizontal shortening between the TIbetan Plateau using interseismic GPS data collected prior to the Wenchuan earthquake.
Next, we show how this geometry can drives the steep velocity gradients far to the northwest of the range front, thus explaining the absence of a range-front interseismic velocity gradient.
Finally, we discuss the effects of Longmen Shan fault geometry on regional seismic hazard and tectonics.

@@

Our fault geometry is derived from a Community Fault Model for the Longmen Shan and Sichuan Basin region being actively developed (Plesch and Shaw, personal communication).
This geometry is derived from structural interpretation of both seismic reflection surverys and surface geology (Hubbard \& Shaw).
We make no modifications to the geometry except to take a fault-perpindicular cross-section.
The cross-section and geometry with depth contours are shown in Figure \ref{fig:regional_map}.

Because our model includes a deep detachment, we must include interseismic GPS data deep into the interior of the Tibetan Plateau.
We study GPS observations collected prior to the Wenchuan earthquake by (CITES Apel, Banerjee, Calais, Gan, Vigny), assembled into a single reference frame by Loveless and Meade 2011.
These velocities are shown in Figure \ref{fig:regional_map} along with the site of the Wenchuan earthquake and other regional faults.
We include in our estimation velocities that are within the potential far-field influence of a deep detachment under the easternmost Tibetan Plateau and the Longmen Shan.
We exclude velocities that are near major faults like the Xianshuihe fault or the Kunlun fault. 
Later, we demonstrate that our results are reasonably robust to the specific data set that is studied.
Addition of more recent observations could bias our results due to possible postseismic transient deformation.

To analyze these data, we use a new quasistatic elastic boundary element code.
An important feature of our code is the inclusion of accurate surface topography in steep mountain ranges or Earth curvature in large-scale problems.
In addition, we can accurately reproduce half-space solutions to very high precision. 
The steep topography of the Longmen Shan range front is included via the SRTM-30 dataset (CITE SANDWELL).

>>>>

\title{Rapid shortening at the eastern margin of the Tibetan plateau prior to the 2008 Mw=7.9 Wenchuan earthquake}
\author{T. Ben Thompson, Brendan J. Meade}
\date{September 2014}
\usepackage{graphicx}
\usepackage[margin=1.0in]{geometry}
\usepackage[nomarkers,figuresonly]{endfloat}

@@

However, it remains to be explained how slow present-day geodetic shortening estimates can be reconciled with the Wenchuan earthquake and the large structurally-observed shortening. Prior estimates can be categorized into those that neglect earthquake cycle effects, analyzing the fault slip-deficit without no explicit model of the geometry or tectonics, (CHEN, SHEN) and those that use block models, which normally assume a simplified fault geometry (CITE MEADE 2007, BURCHFIEL 2008, LOVELESS AND MEADE 2011). Loveless and Meade report the highest shortening rate of 3.2 mm/yr.  Here, we present a model that includes earthquake cycle effects to explain the regional geodetic velocities as the result of a complex interseismically locked fault geometry, including a 20km deep detachment and the range-front thrust Beichuan fault.  This geometry is consistent with structural interpretations and with the geometry of Qi et al 2012 and Fielding et al 2013.  Using this range-front thrust plus detachment geometry, we infer an average horizontal shortening between the TIbetan Plateau using interseismic GPS data collected prior to the Wenchuan earthquake.  Next, we show how this geometry can drive the steep velocity gradients far to the northwest of the range front, thus explaining the absence of a range-front interseismic velocity gradient. Importantly, this model provides increased estimates of seismic hazard. This model proposes that strain accumulates deep in the interior of the Tibetan plateau while being released at the Longmen Shan range front, thus providing a unified theory of the Longmen Shan earthquake cycle.

@@

The fault geometry we consider is derived from a Community Fault Model for the Longmen Shan and Sichuan Basin region being actively developed (Plesch and Shaw, personal communication).  This geometry is derived from structural interpretation of both seismic reflection surverys and surface geology (Hubbard \& Shaw).  We make no modifications to the geometry except to take a fault-perpindicular cross-section.  The cross-section and geometry with depth contours are shown in Figure \ref{fig:regional_map}.

Because this model includes a deep detachment, we must include interseismic GPS data deep into the interior of the Tibetan Plateau. We study GPS observations collected prior to the Wenchuan earthquake by (CITES Apel, Banerjee, Calais, Gan, Vigny), assembled into a single reference frame by Loveless and Meade 2011. These velocities are shown in Figure \ref{fig:regional_map} along with the site of the Wenchuan earthquake and other regional faults. We include in the estimation velocities that are within the potential far-field influence of a deep detachment under the easternmost Tibetan Plateau. We exclude velocities that are near major faults like the Xianshuihe fault or the Kunlun fault. Later, we demonstrate that our results are reasonably robust to the specific data set that is studied. Addition of more recent observations could bias our results due to possible postseismic transient deformation.

To analyze these data, we use new quasistatic elastic boundary element software. An important feature of our tool is the inclusion of accurate surface topography in steep mountain ranges or Earth curvature in large-scale problems. In addition, we can accurately reproduce half-space solutions to very high precision. The steep topography of the Longmen Shan range front is included via the SRTM-30 dataset (CITE SANDWELL).

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\linewidth]{vels_geom_interior.pdf}
    \caption{Vels_geom_interior}
    \label{fig:vels_geom_interior}
\end{figure}
In

A Beichuan-only fault geometry would predict very little deformation beyond 100km (CITE savage 1983), leading one to exclude velocities beyond this range. Both the observed and predicted profile-parallel velocities shown in Figure \ref{fig:big_stack} make clear that such an estimate would result in a near-zero shortening estimate.
